---
title: "Visualization of NYT COVID-19 Data"
output: 
  html_document:
    includes:
       in_header: ga_script.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### These visualizations update daily. [Link to data here](https://github.com/nytimes/covid-19-data).
#### Last updated `r format(Sys.time(), "%B %d, %Y")` at `r format(Sys.time(), "%H:%M %Z")`. 


A compliment to the [map created by the New York Times](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html) showing the spread of COVID-19 cases by U.S. County created by [Stan Oklobdzija](stanokl.github.io). Unlike the Times' map, this map is normalized by county population size to show the penetration of COVID-19 cases. Counties are shaded based on the number of COVID-19 cases per 1,000 residents. Information on the raw number cases and deaths, the rates per 1,000 residents and the [doubling time](https://en.wikipedia.org/wiki/Doubling_time) are also available by clicking on the county. 


A note on the data: The New York Times aggregated all five boroughs of New York City into one geographic entity. To map it, I condensed data on that city to the borough of Manhattan, (New York County.) Similarly, the Times created a separate geographic entity for Kansas City, Missouri. I reaggregated those cases to Jackson County, Missouri which houses that city. 


```{r, echo = FALSE, message=FALSE, warning=FALSE, comment=NA, out.width='100%'}
library(tidyverse)
library(tidycensus)
library(htmltools)

##NYT data
nyt_data <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv") %>% 
  mutate(fips=str_pad(fips, width=5,side="left",pad="0")) %>%
  mutate(fips=ifelse(county=="New York City", "NYC",fips)) %>%
  mutate(fips = ifelse(county=="Kansas City", "29101",fips)) %>% group_by(fips, date) %>%
  mutate(cases=sum(cases), deaths=sum(deaths)) %>% distinct(date,fips, .keep_all = T)

  
##load census data
##I initially loaded several county level variables, but in the end I just used population

# key <- "KEY GOES HERE"
# 
# census_api_key(key, install=T)
# #
# # variables <- load_variables(year = 2018, dataset= "acs5", cache = TRUE)
# # # #
# pop_county <- get_acs(geography = "county",
#                       variables = "B01003_001") %>% mutate("population"=estimate) %>%
#   select(-c(variable,estimate))
# 
# med_inc <- get_acs(geography = "county",
#                    variables = "B19013_001") %>% mutate("med_income"=estimate) %>%
#   select(-c(variable,estimate,moe))
# 
# med_age <- get_acs(geography = "county",
#                    variables = "B01002_001") %>% mutate("med_age"=estimate)%>%
#   select(-c(variable,estimate,moe))
# 
# whites <- get_acs(geography = "county",
#                   variables = c("B03002_003","B02001_001")) %>%
#   group_by(GEOID, NAME) %>% summarise(white_pct=estimate[variable=="B03002_003"]/estimate[variable=="B02001_001"])
# 
# county_demos <- merge(pop_county, med_inc, by=c("GEOID","NAME")) %>%
#   merge(med_age, by=c("GEOID","NAME")) %>% merge(whites, by=c("GEOID","NAME"))
# #
# # ##make an NYC line for NYT data
# #
# nyc <- county_demos %>% filter(NAME %in% c("New York County, New York",
#                                            "Kings County, New York",
#                                            "Queens County, New York",
#                                            "Bronx County, New York",
#                                            "Richmond County, New York")) %>%
#   summarise(population=sum(population), med_income=mean(med_income),med_age=mean(med_age), white_pct=mean(white_pct)) %>%
#   mutate(GEOID="NYC") %>% slice(rep(1,each=5))
# #
# # # nyc$GEOID <- county_demos$GEOID[county_demos$NAME %in% c("New York County, New York",
# # #                                                          "Kings County, New York",
# # #                                                          "Queens County, New York",
# # #                                                          "Bronx County, New York",
# # #                                                          "Richmond County, New York")]
# #
#  nyc$GEOID <- "NYC"
# #
# # # #
# # # #
# kc_whites <-  get_acs(geography = "place", variables = c("B03002_003","B02001_001")) %>%
#   filter(NAME=="Kansas City city, Missouri") %>%
#   group_by(GEOID, NAME) %>% summarise(white_pct=estimate[variable=="B03002_003"]/estimate[variable=="B02001_001"])
# 
# kc_whites$GEOID <- "kcmo"
# 
# kc_mo <-get_acs(geography = "place", variables = c("B01003_001","B01002_001","B19013_001")) %>%
#   filter(NAME=="Kansas City city, Missouri") %>% spread(variable, estimate) %>%
#   mutate(GEOID="kcmo") %>%
#   rename(med_age = "B01002_001", population = "B01003_001", med_income = "B19013_001") %>%
#   select(-moe) %>%group_by(GEOID) %>% summarise_all(funs( na.omit(unique(.)) ))
# 
# kc_mo <- merge(kc_whites, kc_mo, by=c("GEOID","NAME"))
# 
# kc_mo$GEOID <- "29101"
# # #
#  county_demos <- bind_rows(county_demos, nyc, kc_mo)
# # #
#  save(county_demos, file="YOUR PATH HERE")

load("PATH TO YOUR COUNTY DEMO DATA")

##merge county_demo and nyt_data
##I calculated several statistics, but in the end just ended up using growth rates in the visualizations. 
library(lubridate)

nyt_demo <- merge(nyt_data, county_demos, by.x="fips",by.y="GEOID") %>% select(-moe) %>%
  group_by(fips) %>% mutate(case_per_1k = (cases/population) * 1000, death_per_1k=(deaths/population) *1000) %>%
  mutate(growth_rate_5day = ((cases-lag(cases, n=5, order_by = date))/lag(cases, n=5, order_by = date))*100,
    case_p1k_grow1 = case_per_1k-lag(case_per_1k, n=1, order_by = date),
         case_p1k_grow3 = case_per_1k-lag(case_per_1k, n=3, order_by = date),
         case_p1k_grow5 = case_per_1k-lag(case_per_1k, n=5, order_by = date),
         death_p1k_grow1 = death_per_1k-lag(death_per_1k, n=1, order_by = date),
         death_p1k_grow3 = death_per_1k-lag(death_per_1k, n=3, order_by = date),
         death_p1k_grow5 = death_per_1k-lag(death_per_1k, n=5, order_by = date),
    growth_rate = (log(cases)-lag(log(cases), n=1, order_by=date)),
    double_time = log(2)/growth_rate) %>%
  mutate(month=month(date), date=ymd(date)) %>%
  rename(GEOID=fips) 

##recode NYC data to manhattan
  
nyt_demo$GEOID <- ifelse(nyt_demo$GEOID=="NYC", 36061, nyt_demo$GEOID)

nyt_demo$NAME <- ifelse(nyt_demo$GEOID=="36061", "NYC (all boroughs)",nyt_demo$NAME)


### create interactive map
library(tigris)
library(leaflet)

map_shape <- counties(cb=T, year=2018)

map_data <- nyt_demo %>% group_by(GEOID) %>% filter(date==max(date)) %>%
  mutate(cases_per_1k = (cases/population) * 1000,
         deaths_per_1k = (deaths/population) * 1000) %>%
  distinct(GEOID, .keep_all=T) %>% mutate(double_time = ifelse(double_time==Inf, NA,double_time))

leafmap <- geo_join(map_shape, map_data, by="GEOID")

##create function to compute percentile rank

perc.rank <- function(val, data)  length(data[data <= val])/length(data)*100

popup_dat <- paste0("<strong>County: </strong>", 
                    leafmap$NAME.1, 
                    "<br><strong>Cumulative Cases: </strong>", 
                    prettyNum(leafmap$cases,big.mark = ","),
                    "<br><strong>Cumulative Deaths: </strong>", 
                    prettyNum(leafmap$deaths, big.mark=","),
                    "<br><strong>Cumulative Cases per 1k: </strong>", 
                    round(leafmap$cases_per_1k, digits=3),
                    "<br><strong>Cumulative Deaths per 1k: </strong>", 
                    round(leafmap$deaths_per_1k, digits=3),
                    "<br><strong>Daily Case Growth Rate: </strong>", 
                    round(leafmap$growth_rate * 100, digits=2))

##define break points
breaks <- c(0, .1,.3,.5,.7,1,2,4,8,max(map_data$cases_per_1k))

pal <- colorBin("Reds",domain=leafmap$cases_per_1k, na.color="grey",bin=breaks)

# Render final map in leaflet.
#library(htmlwidgets)

covid_map <-leaflet(data = leafmap, width = "100%") %>% addTiles() %>%
  setView(lat=39.8283,lng= -98.5795, zoom=4) %>%
  addPolygons(fillColor = ~pal(cases_per_1k), 
              fillOpacity = 0.8, 
              color = "#BDBDC3", 
              weight = 1,
              popup = popup_dat) %>%
    addLegend(pal = pal, 
            values = ~cases_per_1k, 
            position = "bottomright", 
            title = "Cases Per 1,000 Residents:")

covid_map

```



#### Top 30 Counties by Daily Growth Rate, (for counties with more than 100 cases)


```{r, echo = FALSE, message=FALSE, warning=FALSE, comment=NA, out.width='100%'}
library(plotly)
library(ggrepel)

fast_counties <- nyt_demo %>% ungroup() %>% filter(date==max(date) & cases > 100) %>% distinct(GEOID, .keep_all=T) %>% top_n(30, growth_rate) %>% arrange(-growth_rate)


f <- list(
  size = 18,
  color = "#7f7f7f"
)
x <- list(
  title = "Cases",
  titlefont = f
)
y <- list(
  title = "Growth Rate",
  titlefont = f
)


plot_ly(fast_counties, x=~cases, y=~growth_rate, size=~population, color=~state, type="scatter", mode="markers",
  hoverinfo="text", 
         text = ~paste('</br> Name: ', NAME,
                       '</br> Cases: ', prettyNum(cases, big.mark = ","),
                       '</br> Population: ', prettyNum(population, big.mark = ","),
                       '</br> Daily Growth Rate: ', round(growth_rate * 100, digits=2)
                       )) %>% layout(showlegend = FALSE, yaxis = list(range=c(0, max(fast_counties$growth_rate)+.2))) %>% layout(xaxis = x, yaxis=y) %>% config(displayModeBar = FALSE)

```


#### Doubling Time in Largest U.S. Metro Areas.
(Doubling time calculated based on average growth rate of previous week.)


```{r, echo = FALSE, message=FALSE, warning=FALSE, comment=NA, out.width='100%', out.height='100%'}
library(tigris)
library(plotly)
library(ggrepel)
library(readxl)
library(lubridate)
library(tidyverse)

##reload Times Data
##mutate NYC to manhattan and KC to Johnson County

nyt_data <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv") %>% 
  mutate(fips=str_pad(fips, width=5,side="left",pad="0")) %>%
  mutate(fips=ifelse(county=="New York City", "36061",fips)) %>%
  mutate(fips = ifelse(county=="Kansas City", "29101",fips)) %>%
  mutate(date = ymd(date))
  

##Metro County List

##counties that encompass each Census CSA area found here: 
##https://www2.census.gov/programs-surveys/metro-micro/geographies/reference-files/2018/delineation-files/list1_Sep_2018.xls

metros <- read_xls("list1_Sep_2018.xls", skip = 2)

##names of 30 most populous metro areas in US
##https://www.thoughtco.com/largest-metropolitan-areas-1435135
##Note: amend some of the names that don't match the census CSA names

top_metros <- read.csv("top_metros.csv", stringsAsFactors = F)

metros <- metros[metros$`CSA Title` %in% top_metros$name,] 

metros <- metros %>% mutate(fips=paste0(metros$`FIPS State Code`,metros$`FIPS County Code`)) %>%
  merge(nyt_data, by="fips")

metro_sum <- metros %>% group_by(date, `CSA Title`) %>% summarise(cases=sum(cases)) %>% group_by(`CSA Title`) %>%
  mutate(growth_rate = log(cases)-lag(log(cases), n=1, order_by=date)) %>%
  group_by(week(date), `CSA Title`) %>% mutate(wk_growth_rate = mean(growth_rate, na.rm=T), 
         double_time = log(2)/wk_growth_rate) %>%
  mutate(double_time = ifelse(double_time==Inf,NA,double_time))


f <- list(
  size = 18,
  color = "#7f7f7f"
)
x <- list(
  title = "Metro Area",
  titlefont = f
)
y <- list(
  title = "Doubling Time (Days)",
  titlefont = f
)

for_plot <- metro_sum[metro_sum$date==max(metro_sum$date),]

for_plot$`CSA Title` <- factor(for_plot$`CSA Title`, levels = unique(for_plot$`CSA Title`)[order(for_plot$double_time, decreasing = TRUE)])

plot_ly(for_plot, y=~double_time, x=~as.factor(`CSA Title`),color=~`CSA Title`, type="bar",
  hoverinfo="text", 
         text = ~paste('</br> Name: ', `CSA Title`,
                       '</br> Cases: ', prettyNum(cases, big.mark = ","),
                       '</br> Doubling Time: ', paste0(prettyNum(double_time, digits=3), " days")
                       )) %>% layout(showlegend = FALSE) %>% layout(xaxis = x, yaxis=y, autosize=F, height=800, width=1200) %>% config(displayModeBar = FALSE)

```
